sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-02 doc/paper-jss/owls.Rmd
Editing

```
#>   Non-linear minimizer suggests successful convergence
#>   Hessian matrix is positive definite
#>   No extreme or very small eigen values detected
#> x `ln_tau_E` gradient > 0.001
#>   See `?run_extra_optimization()`
#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
#> 
#> x `ln_phi` gradient > 0.001
#>   See `?run_extra_optimization()`
#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
#> 
#>   No fixed-effect standard errors are NA
#>   No fixed-effect standard errors look unreasonably large
#>   No sigma parameters are < 0.01
#>   No sigma parameters are > 100
#>   Range parameter doesn't look unreasonably large
```

@@

```{r owl-newton, results='hide', eval=!rds_saved2}
fit_owl2 <- run_extra_optimization(fit_owl, nlminb_loops = 0, newton_loops = 1)
```

@@

```
#> Non-linear minimizer suggests successful convergence
#> Hessian matrix is positive definite
#> No extreme or very small eigen values detected
#> No gradients with respect to fixed effects are >= 0.001
#> No fixed-effect standard errors are NA
#> No fixed-effect standard errors look unreasonably large
#> No sigma parameters are < 0.01
#> No sigma parameters are > 100
#> Range parameter doesn't look unreasonably large
```

@@

We can also test for zero-inflation using the `simulate.sdmTMB()` function and a couple functions from the \pkg{DHARMa} package.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.

```{r zero-test-owls, echo=TRUE, message=FALSE}
s_nb2 <- simulate(fit_owl2, nsim = 500)
pred_fixed <- fit_owl2$family$linkinv(pred$est_non_rf)
r_nb2 <- DHARMa::createDHARMa(
  simulatedResponse = s_nb2,
  observedResponse = fit_owl2$data$count,
  fittedPredictedResponse = pred_fixed
)
DHARMa::testZeroInflation(r_nb2, plot = FALSE)
```

@@

p_mean <- p %>%
  group_by(CBCID) %>% summarise(
    year = max(year),
    X = mean(X), Y = mean(Y),
    mean_est_count = exp(mean(est)),
    nao_effect = b[b$term == "nao", "estimate"] + mean(zeta_s),
    nao_sim = mean(nao_effect_sim),
    nao_lwr = mean(nao_effect_lwr),
    nao_upr = mean(nao_effect_upr)
  )
```

>>>>

```
#> ✓ Non-linear minimizer suggests successful convergence
#> ✓ Hessian matrix is positive definite
#> ✓ No extreme or very small eigenvalues detected
#> ✗ `ln_tau_E` gradient > 0.001
#>   See `?run_extra_optimization()`
#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
#> 
#> ✗ `ln_phi` gradient > 0.001
#>   See `?run_extra_optimization()`
#>   Or refit with `control = sdmTMBcontrol(newton_loops = 1)`
#> 
#> ✓ No fixed-effect standard errors are NA
#> ✓ No fixed-effect standard errors look unreasonably large
#> ✓ No sigma parameters are < 0.01
#> ✓ No sigma parameters are > 100
#> ✓ Range parameter doesn't look unreasonably large
```

@@

```{r owl-newton, results='hide', eval=!rds_saved2}
fit_owl2 <- run_extra_optimization(fit_owl, newton_loops = 1)
```

@@

```
#> ✓ Non-linear minimizer suggests successful convergence
#> ✓ Hessian matrix is positive definite
#> ✓ No extreme or very small eigenvalues detected
#> ✓ No gradients with respect to fixed effects are >= 0.001
#> ✓ No fixed-effect standard errors are NA
#> ✓ No fixed-effect standard errors look unreasonably large
#> ✓ No sigma parameters are < 0.01
#> ✓ No sigma parameters are > 100
#> ✓ Range parameter doesn't look unreasonably large
```

@@

We can also test for zero-inflation using the `simulate.sdmTMB()` function.
If our data were zero-inflated or overdispersed, we might have needed to use a delta ("hurdle") model instead (e.g., `delta_truncated_nbinom2()` or possibly even `delta_gamma()` if we were willing to treat the counts as continuous.

```{r zero-test-owls, echo=TRUE, message=FALSE}
s_nb2 <- simulate(fit_owl2, nsim = 400)
```

The proportion of zeros looks approximately the same in the simulated and real data.

```{r zero-test-owls2, echo=TRUE}
mean(s_nb2 == 0)
mean(snow$count == 0)
```

And we can more formally test this with functions from the \pkg{DHARMa} package.

```{r zero-test-owls3, echo=TRUE}
pred_fixed <- fit_owl2$family$linkinv(pred$est_non_rf)
r_nb2 <- DHARMa::createDHARMa(
  simulatedResponse = s_nb2,
  observedResponse = fit_owl2$data$count,
  fittedPredictedResponse = pred_fixed
)
DHARMa::testZeroInflation(r_nb2, plot = FALSE)
```

@@

p_mean <- p %>%
  group_by(CBCID) %>% summarise(
    year = max(year),
    X = mean(X), Y = mean(Y),
    mean_est_count = exp(mean(est)),
    nao_effect = b$estimate[b$term == "nao"] + mean(zeta_s),
    nao_sim = mean(nao_effect_sim),
    nao_lwr = mean(nao_effect_lwr),
    nao_upr = mean(nao_effect_upr)
  )
```

### 2022-11-02 doc/paper-jss/paper-jss.Rmd
Editing

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

@@

\begin{landscape}
\begin{figure}[p]
\centering 
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix \ref{app:model} for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} [@wood2017a], \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates.
It usually makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `sdmTMB::add_utm_columns()`].
Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial coordinates `X` and `Y`, which are UTM coordinates in kilometres.

@@

```{r sdmTMB-lib, eval=TRUE}
library(sdmTMB)
```

```{r, echo=TRUE, eval=FALSE}
library(sdmTMB)
head(pcod)
```

```{}
#>    year density depth     X     Y
#>   <int>   <dbl> <dbl> <dbl> <dbl>
#> 1  2003   113.    201  446. 5793.
#> 2  2003    41.7   212  446. 5800.
#> 3  2003     0     220  449. 5802.
```

We start by creating a mesh object that contains matrices to apply the SPDE approach.

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
Alternatively, we could have created any mesh via the \pkg{INLA} package and supplied it to `make_mesh()`.
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \@ref(fig:pcod-fig)A).

@@

The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} (REF) or \pkg{broom.mixed} (REF) packages:

@@

```
#> ✓ Non-linear minimizer suggests successful convergence
#> ✓ Hessian matrix is positive definite
#> ✓ No extreme or very small eigen values detected
#> ✓ No gradients with respect to fixed effects are >= 0.001
#> ✓ No fixed-effect standard errors are NA
#> ✓ No fixed-effect standard errors look unreasonably large
#> ✓ No sigma parameters are < 0.01
#> ✓ No sigma parameters are > 100
#> ✓ Range parameter doesn't look unreasonably large
```

@@

@@

The output of `predict()` is a data frame containing overall estimates in link space, estimates from the non-random-field components (intercept and depth), and estimates from the random field components.
We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig.\ \ref{fig:pcod-fig}B--D.

@@

<!-- This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop.  -->
We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig.\ \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \@ref(fig:timing), Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->

>>>>

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>",
  prompt = FALSE
  # R.options = list(prompt = " ", continue = " ")
)
```

@@

\begin{landscape}
\begin{figure}[p]
\centering 
\includegraphics[width=9in]{../../figs/diagram}
\caption[]{
Components of an \pkg{sdmTMB} model with illustrations, descriptions, examples, notation, and example code.
An \pkg{sdmTMB} model can be built from any combination of the process components (first six rows) plus an observation component (last row).
The examples are from an SDM context, but the model can be fit to any spatially referenced point data.
Notation:
We refer to design matrices as $\bm{X}$.
The indexes $\bm{s}$, $t$, and $g$ index spatial coordinates, time, and group, respectively.
The $\sigma$ and $\bm{\Sigma}$ symbols represent standard deviations and covariance matrices, respectively.
All other symbols refer to the  described model components (e.g., $\bm{\beta}$ and $\bm{\omega}$ refer to a vector of main effects and spatial random field deviations, respectively).
See Appendix \ref{app:model} for a full description of the model.
Note that \texttt{s()} denotes a smoother as in \pkg{mgcv} \citep{wood2017a}, \texttt{breakpt()} denotes a breakpoint 'hockey-stick' shape \citep[e.g.][]{barrowman2000}, \texttt{(1|g)} denotes a random intercept by group \texttt{g}, and \texttt{~ 0} is used in an \proglang{R} formula to omit an intercept.}\label{fig:diagram}
\end{figure}
\end{landscape}

@@

Here, we illustrate a spatial model fit to Pacific cod (*Gadus macrocephalus*) trawl survey data from Queen Charlotte Sound, British Columbia, Canada.
Our model contains a main effect of depth as a penalized smoother, a spatial random field, and Tweedie observation error.
Our data frame `pcod` (built into the package) has a column `year` for the year of the survey, `density` for biomass density of Pacific cod in the area swept for a given survey tow, `depth` for depth in meters of that tow, and spatial longitude and latitude `lon` and `lat`.

@@

```{r sdmTMB-lib0, eval=TRUE, echo=FALSE}
library(sdmTMB)
# simplify for paper:
pcod <- dplyr::select(pcod, year, density, depth, lon, lat) |> 
  tibble::as_tibble()
```

```{r sdmTMB-lib, eval=TRUE, echo=TRUE}
library(sdmTMB)
head(pcod, n = 3)
```

<!-- An \pkg{sdmTMB} model requires a data frame that contains a response column, columns for any predictors, and columns for spatial coordinates. -->
Usually it makes sense to convert the spatial coordinates to an equidistant projection such as the Universal Transverse Mercator (UTM) to ensure that distance remains constant throughout the study region [e.g., using `sf::st_transform()`, @pebesma2018, or `add_utm_columns()`].
Here we use `add_utm_columns()` to add UTM coordinates with km units (so our estimated spatial range parameter is not too big or small).
By default, the function will guess our UTM zone and create new columns `X` and `Y`.

```{r pcod-utms-eval, echo=FALSE, eval=TRUE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
```

```{r pcod-utms-echo, echo=TRUE, eval=FALSE}
pcod <- add_utm_columns(pcod, c("lon", "lat"), units = "km")
#> Detected UTM zone 9N; CRS = 32609.
#> Visit https://epsg.io/32609 to verify.
```

```{r pcod-head2, echo=TRUE}
head(pcod, n = 3)
```

We then create a mesh object that contains triangulation and projection matrices needed to apply the SPDE approach.

@@

Here, `cutoff` defines the minimum allowed distance between mesh vertices in the units of `X` and `Y`.
`make_mesh()` is a wrapper function for `INLA::inla.mesh.create()`.
Alternatively, we could have created any mesh via the \pkg{INLA} package and supplied it to the `mesh` argument in `make_mesh()`.
We can inspect our mesh object with the associated plotting method (`plot(mesh)`; see Fig. \ref{fig:pcod-fig}A).

@@

The output indicates our model was fit by maximum (marginal) likelihood (`ML`).
We also see the formula, mesh, fitted data, and family.
Next we see any estimated main effects including the linear component of the smoother (`sdepth`), the standard deviation on the smoother weights (`sds(depth)`), the Tweedie dispersion and power parameters, the Matérn range distance, the marginal spatial field standard deviation, and the negative log likelihood at convergence.
The smoother summary follows the format used in the \pkg{brms} package [@brms].
As noted in the summary method, we can extract a tidy data frame similar to using the \pkg{broom} [@broom] or \pkg{broom.mixed} [@broommixed] packages:

@@

```
#> ✓ Non-linear minimizer suggests successful convergence
#> ✓ Hessian matrix is positive definite
#> ✓ No extreme or very small eigenvalues detected
#> ✓ No gradients with respect to fixed effects are >= 0.001
#> ✓ No fixed-effect standard errors are NA
#> ✓ No fixed-effect standard errors look unreasonably large
#> ✓ No sigma parameters are < 0.01
#> ✓ No sigma parameters are > 100
#> ✓ Range parameter doesn't look unreasonably large
```

@@

```{r qcs-head, echo=TRUE}
head(qcs_grid, n = 3)
```

@@

The output of `predict()` is a data frame containing overall estimates in link space (`est`), estimates from the non-random-field components (intercept and depth) (`est_non_rf`), and estimates from the random field components combined (`est_rf`), and the individual random field components (here `omega_s`---the spatial field).

```{r pcod-p-tibble}
p <- tibble::as_tibble(p)
```

```{r pcod-head-predict, echo=TRUE}
head(p, n = 3)
```

We show a basic plot of the estimated spatial random field, predictions across a depth gradient, and predictions in space in Fig. \ref{fig:pcod-fig}B--D.

@@

<!-- This model took approximately `r round(as.numeric(t2 - t1), 0)` seconds to fit on our laptop.  -->
We could inspect the anisotropy with `plot_anisotropy()`, which plots an ellipse showing the spatial range with the spatial coordinates centered on zero.
We could generate an area-weighted population index (e.g., a relative or absolute index of abundance or biomass) that is independent of sampling locations by predicting from that fit on a grid covering the area of interest and summing the predicted biomass with the `get_index()` function (Fig. \ref{fig:pcod-fig}E, Appendix \ref{app:pcod}).

@@

We ran a simple speed comparison between \pkg{sdmTMB}, \pkg{INLA}/\pkg{inlabru}, and \pkg{mgcv} for fitting an SPDE spatial random field model to 1000 data points with Gaussian error across a range of mesh resolutions (Fig. \ref{fig:timing}, Appendix \ref{app:speed}).
In this test, \pkg{sdmTMB} was the fastest across all mesh resolutions; however, \pkg{INLA}/\pkg{inlabru} and \pkg{spaMM} were less affected by mesh resolution than \pkg{sdmTMB}.
\pkg{mgcv} was most affected by mesh resolution.
<!-- With the random effect normalizing constant calculated in R (`control = sdmTMBcontrol(normalize = TRUE)`), \pkg{sdmTMB} was the fastest across all mesh resolutions (this option may not be possible or as fast for all models). -->
<!-- Results with up to an order of magnitude more data slightly shifted the lines up without affecting the qualitative pattern (not shown). -->
<!-- TODO:  Too wordy and informal by describing ‘the lines’. Needs rewording as a discussion about the runtimes. -->
Our test was restricted to one core and default \proglang{R} algebra libraries; all packages could run faster with optimized libraries and parallel processing.
Results with optimized math libraries on one core (openBLAS: @penblas2021 and PARDISO: @pardiso-7.2c) resulted in a ~10% speed increase for \pkg{sdmTMB} and \pkg{inlabru} and a ~7--9-fold speed increase for \pkg{mgcv}.
<!-- Note two other references for PARDISO were listed as required here https://pardiso-project.org/manual/manual.pdf -->
<!-- LB: also worth noting which packages do/do not have parallel infrastructure (notably that \pkg{VAST} does not) -->
